The importance of measurement uncertainty in ecological management

### 2018-01-31 manuscripts/manuscript.Rmd
WIP

abstract: |

  We settle a long standing paradox in ecological management regarding the influence of measurement uncertainty.  Previous work has frequently suggested that the optimal management of fisheries should respond to increased measurement uncertainty by increasing harvest effort, contrary to intuition and the precautionary principle. Through the use of cutting-edge point-based partially observed Markov decision process (POMDP) methods, we demonstrate how this paradox arises from flaws in the standard management theory, which contributes to over-exploitation of fisheries and increased probability of economic and ecological collapse.  In contrast, we find POMDP-based management avoids such over-exploitation while also generating higher economic value through adaptive learning.

@@

However, it is important to remember that both existing methods we discuss and the POMDP approach proposed here can or are already applied to more complex models as well, including those with age or stage structure.  Likewise, while we consider the model and it's parameters as "given," it should be understood to mean that this in practice comes as the result of some model choice and statistical estimation from historical data, frequently accounting for uncertainty in both the process and the measurements [e.g. @Dichmont2016].  As we shall see here, having included uncertainty in the model *estimation* in no way excuses us from having to also deal once again with that uncertainty in the decision process.

@@

The optimization problem is to select the action $a_t$ that will maximize the net present utility over all time. Future utility may be discounted by a factor $\gamma$, so that a value  $V$ in $t$ year is valued at $\gamma^t V$ today.  Numerically, each of these functions are defined over a discrete set of possible states, observations, and actions, and can thus be represented as a collection of matrices or tensors.  While the solution method for MDP problems, stochastic dynamic programming (SDP) is well known and readily implemented in ecological problems [e.g @Mangel1988; @Marescot2013], the POMDP problem cannot be efficiently solved using SDP. While not unknown to the conservation literature [@Williams2011], algorithms for POMDP have historically scaled quite poorly, and their application has been restricted to contexts with only a handful of possible states and actions [e.g. @Chades2008; @Chades2011; @Fackler2014; @Fackler2014b].  POMDP algorithms remain a very active area in artificial intelligence [@Smallwood1973; @Sondik1978; @Pineau2003; @Kaelbling1998], and by adapting recent algorithmic developments [@Kurniawati2008] from that field we are able to find solutions for the considerably more complex problems such as the case of fisheries management, which requires on the order of 100 states and actions to provide sufficient numerical resolution. Readers interested in the technical details of these algorithms should consult the extensive literature in in POMDP methods. For those interested in exploring or applying these methods, we have developed a preliminary R package [@sarsop-pkg], which can be used to replicate any of the results presented here and explore other variations.

@@

Different attention to prior events fundamentally drives the differences between the POMDP strategy and the other strategies and can resolve the uncertainty paradox.  While the catch quota under both TAC and CE strategies can be completely determined given the most recent observation of the stock size by using the policy curves shown in Figure \ref{policy}, this is not the case for the POMDP approach. This fundamental difference is key to understanding the difference in performance and resolving the paradox of uncertainty.  The POMDP policy cannot be specified by the most recent observation alone. Instead, the POMDP policy depends on all prior observations, not just the most recent.  The reason for this complexity comes from the Markov property.  Observations of the state in the perfectly observed system satisfy the Markov property: once we have measured the current biomass exactly, we cannot get any better estimate of the current stock size by studying older measurements. When measurements are uncertain this is no longer the case: intuitively, by comparing the most recent measurement to previous observations we may be able to infer when any given measurement is unusually high or unusually low.  POMDP formalizes this intuition by capturing the information from all previous observations into a *prior belief*.  This prior belief is updated after every subsequent action and observation in accordance with Bayes Law.  The mechanics of this process are well documented in the extensive literature on POMDPs [e.g. @Smallwood1973; @Sondik1978; @Kurniawati2008; @Williams2011], but for our purposes it is sufficient to observe how this evolving prior belief serves to continually adjust the POMDP policy.

@@

**A resolution to the paradox**
Herein lies our resolution to the paradox of uncertainty.  Previous work created this paradox by suggesting that increased harvest rates (decreased target escapement) would often be the rational response to increased uncertainty.  The exact solutions from the POMDP reveal that this is only an accident of the assumptions: it is indeed true that under certain circumstances, harvest levels should increase relative to the case of no uncertainty, but *only when prior knowledge suggests the stock size should be much higher than the most recent estimate would suggest*.  In the POMDP solution, all information must be put into its historical (and constantly updated) context.  When a measurement roughly matches the expectation of this prior context, Figure \ref{policy} shows that the optimal response from POMDP is roughly comparable or slightly more cautious to the harvest under no uncertainty, and not more aggressive as the paradox would suggest. Measurements that exceed expectations are tempered with some skepticism: while the CE solution is willing to meet a high stock measurement with a large harvest, the POMDP solutions increase harvest more cautiously.

@@

So far we have focused on simulation and an examination of the relative policies under different prior information.  Figure \ref{historical} compares the harvest level that would have been recommended by each of the strategies we have considered here against the historically observed catch recorded for two commercially fished stocks: Argentine Hake and European Plaice.  Historical estimates of biomass and catch are taken from the R.A. Myers Legacy Stock Assessment database [@RAM]. Posterior distributions for parameters for the Graham-Schaefer model are estimated with uncertainty through Markov Chain Monte Carlo using `nimble` [@nimble] on the historical data, illustrating how this process might be done in more complex models as well (details and code in Supplement, Appendix B).  $B_{MSY}$ and corresponding TAC and CE policies are calculated based on posterior mean estimates of the model parameters, along with POMDP solution assuming a 10% measurement error rate.  Though measurement error could be estimated directly from the raw data, this would not reflect the true measurement uncertainty arising from the stock assessment process.  Each strategy is then compared to historical observations to determine a recommended harvest.  As we have seen, the TAC and CE harvest policies are uniquely determined by the observed stock size relative to $B_{MSY}$, but the POMDP policy must be re-calculated each time step to reflect both the prior observation and prior action.

These historical examples provide another useful lens to compare how the different strategies respond in the face of fluctuations in real data, rather than model simulations.  In both stocks, historical catch almost always exceeds that recommended by any of our strategies, falling closer to the MSY value estimated (see appendix).   In Argentine Hake, the persistent declines result in both POMDP and CE strategies quickly closing down the fishery in an attempt to let the stock recover to a more productive target biomass, while TAC persists with merely a reduced harvest.  A small recovery of biomass in 1995 is met by an immediate uptick in harvest under CE, while the POMDP response is more conservative.  The example of European Plaice illustrates more volatility in the stock size, revealing further differences in the strategies.  Despite this volatility, the TAC level remains relatively level across all five decades, rising or dipping only slightly with changes in stock size.  In contrast, the CE harvest tracks this volatility almost in perfect mirror.  Here, the POMDP solution falls in between these extremes, almost mirroring which-ever of the two policies is more conservative: the POMDP solution matches the dips in harvest taken by CE to allow the stock to recover most quickly, but does not track the doubling of harvests recommended by CE in the mid 1980, increasingly only to the more modest harvest recommended by the TAC policy.   Once again, we see that POMDP solution provides a more consistently precautionary policy than either the over-simplified optimal control solution of CE or the more rule-of-thumb approach represented by TAC.  Though we have seen that the POMDP solution need not always be more conservative (lower harvest) than these strategies, it makes use of the all prior observations to tune the level of caution appropriately.

@@

Using modern algorithms we have been able to crack the nut of measurement uncertainty in the optimal management of marine fisheries and debunk long-standing paradoxes that increased measurement uncertainty should be met with larger harvests.  We have demonstrated that both existing optimal control approaches and existing precautionary approaches result in over-exploitation and long term declines of fish stocks, as well as reduced economic value, while in contrast, the POMDP approach is able to deliver both ecological and economic recovery despite uncertainty in measurements.  These results have important implications for both global fisheries and ecological management more generally.  In marine fisheries, evidence of such over-exploitation has already been well documented [@Worm2006; @Worm2009; @Costello2016]. Our results suggest that measurement uncertainty may contribute to this pattern, and that reality of measurement errors may frustrate stock recovery under current best practices such as MSY, TAC, and CE.  More generally, the study and application of strategies such as MSY and CE are in no way limited to the decisions in fisheries quotas, but underpin much of conservation and natural resource management in forests and other terrestrial ecosystems as well [e.g. @Clark2000].  Our results underscore the realization that measurement uncertainty is not negligible.  Neither optimization approaches which ignore uncertainty (like CE) or heuristic approaches to account for that uncertainty (like TAC) are sufficient for this task.  Rather, adaptive approaches such as POMDP, which make use of previous estimates to continually adjust prior beliefs, are required to avoid negative ecological and economic outcomes.  Nevertheless, it may be tempting to see these results as either trivial or impractical, but a closer inspection reveals they are neither.

@@

## Future directions

>>>>

abstract: |
  Ecological management and decision-making typically focus on uncertainty about the future, but surprisingly little is known about how to account for uncertainty of the present: that is, the realities of having only partial or imperfect measurements.  Our two primary paradigms for handling decisions under uncertainty -- the precautionary principle and optimal control -- have so far given entirely controdictory results.  This paradox is best illustrated in the example of fisheries management, where many ideas that guide our thinking about ecological decision making were first developed.  There, we find that simplistic optimal control approaches have repeatedly concluded that a manager should increase catch quotas when faced with greater uncertainty about the standing stock biomass.  Meanwhile, most current best practices take a more precautionary approach, decreasing catch quotas by a fixed amount to account for uncertainty.  Using comparisons to both simulated and historical catch data, we find that neither approach is sufficient to avoid stock collapses under moderate observational uncertainty.  Using cutting-edge point-based partially observed Markov decision process (POMDP) methods, we demonstrate how this paradox arises from flaws in the standard management theory, which contributes to over-exploitation of fisheries and increased probability of economic and ecological collapse.  In contrast, we find POMDP-based management avoids such over-exploitation while also generating higher economic value through adaptive learning.  These results have significant implications for how we handle uncertainty in both fisheries and ecological management more generally.  Dealing with uncertainty in decisions requires approaches which are both adaptive and explicit about uncertainty, such as POMDPs.

@@

It is important to remember that both existing methods we discuss and the POMDP approach proposed here can or are already applied to more complex models as well, including those with age or stage structure.  Likewise, while we consider the model and it's parameters as "given," it should be understood to mean that this in practice comes as the result of some model choice and statistical estimation from historical data, frequently accounting for uncertainty in both the process and the measurements [e.g. @Dichmont2016].  As we shall see here, having included uncertainty in the model *estimation* in no way excuses us from having to also deal once again with that uncertainty in the decision process.

@@

The optimization problem is to select the action $a_t$ that will maximize the net present utility over all time. Future utility may be discounted by a factor $\gamma$, so that a value  $V$ in $t$ year is valued at $\gamma^t V$ today.  Numerically, each of these functions are defined over a discrete set of possible states, observations, and actions, and can thus be represented as a collection of matrices or tensors.  While the solution method for MDP problems, stochastic dynamic programming (SDP) is well known and readily implemented in ecological problems [e.g @Mangel1988; @Marescot2013], the POMDP problem cannot be efficiently solved using SDP. While not unknown to the conservation literature [@Williams2011], algorithms for POMDP have historically scaled quite poorly, and their application has been restricted to contexts with only a handful of possible states and actions [e.g. @Chades2008; @Chades2011; @Fackler2014; @Fackler2014b].  POMDP algorithms have a substantial literature starting almost contemporaneously to the work of @Reed1979 (see foundational papers by @Smallwood1973 and @Sondik1978) and remain an active area in artificial intelligence [e.g. @Kaelbling1998; @Pineau2003]. By adapting recent algorithmic developments [@Kurniawati2008] from that field we are able to find solutions for the considerably more complex problems such as the case of fisheries management, which requires on the order of 100 states and actions to provide sufficient numerical resolution. As the details for solving such problems numerically are already well documented we will not rehash them here, but instead provide a preliminary R package [@sarsop-pkg] which provides an implementation of the SARSOP algorithm and the fisheries POMDP problem, which can be used to replicate any of the results presented here and explore other variations.  Details of the analysis presented here including annotated code to reproduce and further explore all of the following results is provided in the supplementary material.

@@

Different attention to prior events fundamentally drives the differences between the POMDP strategy and the other strategies and can resolve the uncertainty paradox.  While the catch quota under both TAC and CE strategies can be completely determined given the most recent observation of the stock size by using the policy curves shown in Figure \ref{policy}, this is not the case for the POMDP approach. This fundamental difference is key to understanding the difference in performance and resolving the paradox of uncertainty.  The POMDP policy cannot be specified by the most recent observation alone. Instead, the POMDP policy depends on all prior observations, not just the most recent.  The reason for this complexity comes from the Markov property.  Observations of the state in the perfectly observed system satisfy the Markov property: once we have measured the current biomass exactly, we cannot get any better estimate of the current stock size by studying older measurements. When measurements are uncertain this is no longer the case: intuitively, by comparing the most recent measurement to previous observations we may be able to infer when any given measurement is unusually high or unusually low.  POMDP formalizes this intuition by capturing the information from all previous observations into a *prior belief*.  This prior belief is updated after every subsequent action and observation in accordance with Bayes Law.  The mechanics of this process are well documented in the extensive literature on POMDPs [e.g. @Smallwood1973; @Sondik1978; @Kurniawati2008; @Williams2011], but for our purposes it is sufficient to observe how this evolving prior belief serves to continually adjust the POMDP policy.  Figure S8 illustrates how the prior belief evolves in response to subsequent observations and actions over the course of an individual simulation.

@@

**A resolution to the paradox.**

Herein lies our resolution to the paradox of uncertainty.  Previous work created this paradox by suggesting that increased harvest rates (decreased target escapement) would often be the rational response to increased uncertainty.  The exact solutions from the POMDP reveal that this is only an accident of the assumptions: it is indeed true that under certain circumstances, harvest levels should increase relative to the case of no uncertainty, but *only when prior knowledge suggests the stock size should be much higher than the most recent estimate would suggest*.  In the POMDP solution, all information must be put into its historical (and constantly updated) context.  When a measurement roughly matches the expectation of this prior context, Figure \ref{policy} shows that the optimal response from POMDP is roughly comparable or slightly more cautious to the harvest under no uncertainty, and not more aggressive as the paradox would suggest. Measurements that exceed expectations are tempered with some skepticism: while the CE solution is willing to meet a high stock measurement with a large harvest, the POMDP solutions increase harvest more cautiously.

@@

So far we have focused on simulation and an examination of the relative policies under different prior information.  Figure \ref{historical} compares the harvest level that would have been recommended by each of the strategies we have considered here against the historically observed catch recorded for two commercially fished stocks: Argentine Hake and European Plaice.  Historical estimates of biomass and catch are taken from the R.A. Myers Legacy Stock Assessment database [@RAM]. Posterior distributions for parameters for the Gordon-Schaefer model are estimated with uncertainty through Markov Chain Monte Carlo using `nimble` [@nimble] on the historical data, illustrating how this process might be done in more complex models as well (details and code in Supplement, Appendix B).  $B_{MSY}$ and corresponding TAC and CE policies are calculated based on posterior mean estimates of the model parameters, along with POMDP solution assuming a 10% measurement error rate.  Though measurement error could be estimated directly from the raw data, this would not reflect the true measurement uncertainty arising from the stock assessment process.  Each strategy is then compared to historical observations to determine a recommended harvest.  As we have seen, the TAC and CE harvest policies are uniquely determined by the observed stock size relative to $B_{MSY}$, but the POMDP policy must be re-calculated each time step to reflect both the prior observation and prior action.

These historical examples provide another useful lens to compare how the different strategies respond in the face of fluctuations in real data, rather than model simulations.  In both stocks, historical catch almost always exceeds that recommended by any of our strategies, falling closer to the MSY value estimated (see Appendix B).   In Argentine Hake, the persistent declines result in both POMDP and CE strategies quickly closing down the fishery in an attempt to let the stock recover to a more productive target biomass, while TAC persists with merely a reduced harvest.  A small recovery of biomass in 1995 is met by an immediate uptick in harvest under CE, while the POMDP response is more conservative.  The example of European Plaice illustrates more volatility in the stock size, revealing further differences in the strategies.  Despite this volatility, the TAC level remains relatively level across all five decades, rising or dipping only slightly with changes in stock size.  In contrast, the CE harvest tracks this volatility almost in perfect mirror.  Here, the POMDP solution falls in between these extremes, almost mirroring which-ever of the two policies is more conservative: the POMDP solution matches the dips in harvest taken by CE to allow the stock to recover most quickly, but does not track the doubling of harvests recommended by CE in the mid 1980, increasingly only to the more modest harvest recommended by the TAC policy.   Once again, we see that POMDP solution provides a more consistently precautionary policy than either the over-simplified optimal control solution of CE or the more rule-of-thumb approach represented by TAC.  Though we have seen that the POMDP solution need not always be more conservative (lower harvest) than these strategies, it makes use of the all prior observations to tune the level of caution appropriately.

@@

Using modern algorithms we have been able to crack the nut of measurement uncertainty in the optimal management of marine fisheries and debunk long-standing paradoxes that increased measurement uncertainty should be met with larger harvests.  We have demonstrated that both existing optimal control approaches and existing precautionary approaches result in over-exploitation and long term declines of fish stocks, as well as reduced economic value, while in contrast, the POMDP approach is able to deliver both ecological and economic recovery despite uncertainty in measurements.  These results have important implications for both global fisheries and ecological management more generally.  In marine fisheries, evidence of such over-exploitation has already been well documented [@Worm2006; @Worm2009; @Costello2016]. Our results suggest that measurement uncertainty may contribute to this pattern, as such errors lead to over-exploitation under current best practices such as MSY, TAC or CE, which have previously been thought to be sufficient to ensure the recovery of stocks world-wide [@Costello2016].

Our results also have important consequences for ecological management more generally. The study and application of management strategies such as MSY and CE underpin much of conservation and natural resource management in forest management and many other terrestrial ecosystems [e.g. @Clark2000]. Descision making under uncertainty is a central challenge for ecology and conservation biology [@Polasky2011], which we have long attempted to address through either simplified optimal control models or more heuristic approaches such as the precautionary principle and resilience thinking [@Fischer2009]. We have illustrated how both of these approaches can fail to provide an adequate strategy when faced with measurement uncertainty. To avoid this pitfall, a strategy must be dynamic to make best use of the available information without assuming that information is perfect.  The simplisitic optimal control approach of CE fails by placing too much confidence in measurements even when they are significantly at odds with prior expectations. Similarly, the TAC approach also fails due to it's reliance on a simple one-size-fits-all strategy, not reducing catch quotas sufficiently when stocks are too low but also failing to increase them when stocks are higher.   It is precisely this ability to adapt the policy in response to available information, as illustrated in Figure \ref{policy}, that makes the POMDP approach successful. Nevertheless, it may be tempting to see these results as either trivial or impractical, but a closer inspection reveals they are neither.

@@

## Conclusions and Future directions
